Towards Comparable Active Learning

### 2023-05-31 doc/al_benchmark_paper.tex
paper update

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Choosing the Classifier}\label{sec:choosing_the_classifier}
Traditionally, the classifier is chosen per dataset so that it is capable of solving the dataset close to the SOTA performance reported in the literature.
Similar to our hypothesis in section \ref{sec:evaluation} we hypothesize that AL algorithms will perform similarly on small classifiers and more complex ones, so that the overall ranking of algorithms stays the same.
{\color{red} TODO: Transform this into full hypothesis incl. formatting?}\\
On the basis of this hypothesis we opt to use smaller classifiers that still solve the dataset to a reasonable degree.
Smaller classifiers also require fewer labeled datapoints to acquire performance close to the upper bound.
For every dataset the chosen architecture's hyperparameters are optimized by to archive maximum upper bound performance.
For an overview of architectures and hyperparameters please refer to Appendix \ref{app:hyperparameters}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Training the Classifier}\label{sec:training_the_classifier}
Generally, the classification model can be trained in two ways. Either you reset the parameters after each AL iteration and train the classifier from scratch with the updated labeled set $\mathcal{L}^{(i)}$, or you retain the previous state and only fine-tune the classifier on $\mathcal{L}^{(i)}$ for a reduced number of epochs.
In this work we use the fine-tuning method for raw datasets to save computation, while we use the from-scratch training for embedded dataset, since they have very small classifiers and this method generally produces better results.
Our fine-tuning scheme always trains for at least one epoch and employs an aggressive early stopping after that.
The early stopping has patience 0, so it will stop as soon as the validation loss does no longer decrease.
Even though the use of a fully labeled validation set might be regarded as impractical, since such a set will never exist during deployment, we strongly advocate for using it to control the classifier training.
In this work we use the validation set to optimize the hyperparameters of the classifier and reduce overfitting with early stopping the training process in every iteration.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}\label{sec:datasets}
For all our datasets we use the pre-defined train / test splits, if given. 
In the remaining cases, we define test sets upfront and store them into separate files to keep them fixed across all experiments.
The validation set is split during experiment-time and depends on the dataset-seed.\\
\textbf{Tabular:}
We use \textbf{Splice}, \textbf{DNA} and \textbf{USPS} from LibSVMTools \cite{libsvmtools}.
All three datasets are normalized between [0, 1]. \\
\textbf{Image:}
We use \textbf{FashionMNIST} \cite{xiao2017fashion} and \textbf{Cifar10} \cite{krizhevsky2009learning}.
Both datasets are normalized according to their standard protocols. \\
\textbf{Text:}
We use \textbf{News Category} \cite{misra2022news} and \textbf{TopV2} \cite{chen-etal-2020-low-resource}.
For News Category we use  the 15 most common categories as indicated by its Kaggle site.
We additionally drop sentences above 80 words to reduce the needed padding (retaining 99,86\% of the data).
For TopV2, we are only using the "alarm" domain.
Both datasets are encoded with pre-trained GloVe embeddings \cite{pennington2014glove}.
Since neither dataset provided a fixed test set, we randomly split 7000 datapoints into a test set. \\ [1mm]
%
We would like to point out that these datasets can be considered "toy-datasets" and therefore not relevant for practical purposes.
This might be true if we aimed to develop novel classification models on these datasets, however we are solely focused on comparing different AL algorithms in this paper.
Our core assumption is that a well-performing algorithm in our benchmark will also transfer into more practical use-cases. \\ [1mm]
Adapting the experimental setting from \cite{hacohen2022active} we offer all our datasets in the raw setting as well as pre-encoded by a fixed embedding model that was trained by unsupervised contrastive learning. 
The text datasets are an exception, as they are only offered in their encoded form.
The pre-encoded datasets enable us to test our single-sample algorithms on more complex datasets like Cifar10 and FashionMnist without the need of sampling $>2000$ datapoints before we can reach our upper bound performance.
The embedding model was trained with the SimCLR \cite{chen2020simple} algorithm. 
For Cifar10 we adapt the reported hyperparameters from \cite{hacohen2022active} and for the tabular datasets we use random search to optimize the hyperparameters.
The quality of embeddings during pretext training was measured after each epoch by attaching a linear classification head and evaluating this classifier for test accuracy, mirroring our AL setup for embedded datasets.

@@

\begin{table}[]
	\centering
	\begin{tabular}{l|lll}
		& Splice        & DNA           & USPS          \\
		\hline
		Oracle          & 0.830 +- 0.01 & 0.836 +- 0.02 & 0.823 +- 0.01 \\
		SAL & 0.799 +- 0.01 & 0.797 +- 0.03 & 0.809 +- 0.01 \\
		Coreset & 0.800 +- 0.01 & 0.795 +- 0.03 & 0.787 +- 0.02 \\
		TypiClust       & 0.790 +- 0.01 & 0.771 +- 0.04 & 0.761 +- 0.02 \\
		MarginScore     & 0.797 +- 0.02 & 0.795 +- 0.04 & 0.808 +- 0.01 \\
		ShannonEntropy  & 0.799 +- 0.02 & 0.794 +- 0.04 & 0.807 +- 0.01 \\
		RandomAgent     & 0.788 +- 0.01 & 0.765 +- 0.03 & 0.772 +- 0.01 \\
		Badge           & 0.807 +- 0.01 & 0.769 +- 0.06 & 0.797 +- 0.02 \\
		BALD            & 0.811 +- 0.01 & 0.743 +- 0.04 & 0.717 +- 0.05
	\end{tabular}
	\caption{AUC values for all algorithms on the tabular datasets. Higher is better.
	{\color{red}Is the STD even helping us here? Since we repeated 50 times, we can use the values as "true medians"} }
\end{table}
%
\begin{figure}
	\centering
	\includegraphics[width=0.49\linewidth]{img/eval_dna}
	\includegraphics[width=0.49\linewidth]{img/eval_splice.png}
	\caption{Results for all algorithms on Splice and DNA, both from the tabular domain. Even within one domain, the performance of the same algorithm can vary drastically.}
	\label{fig:eval_vector}
\end{figure}
%
\begin{figure}
\centering
	\includegraphics[width=0.49\linewidth]{img/vector.png}
	\includegraphics[width=0.49\linewidth]{img/vector_enc.png}
	\includegraphics[width=0.49\linewidth]{img/text.png}
	\includegraphics[width=0.49\linewidth]{img/all_data.png}
	\caption{Critical Difference Diagram for all algorithms grouped by domain and all domains combined. Ranks are computed based on median AUC for each algorithm and dataset combination. Lower ranks are better.}
	\label{fig:cd_diagrams}
\end{figure}
%
From Fig. \ref{fig:eval_vector} we notice drastically different qualities for the same AL algorithm for different datasets.
We would like to highlight that both datasets are tabular from the medical domain with similar number of features and classes, yet we see that i.e. BALD is the best algorithm for Splice and the worst algorithm for DNA.
These inconsistencies are present between the datasets of all our tested domains, further highlighting the difficulties for comparing AL algorithms in terms of average performance.
In order to provide a meaningful analysis of which algorithm can be expected to perform best on average we ranked the algorithm for each dataset based on their median AUC and displayed these rankings in critical difference diagrams \cite{IsmailFawaz2018deep}.
In Fig. \ref{fig:cd_diagrams} we report the rankings split by domain as well as across all domains (excluding the toydata). \\ [2mm]
%
BALD performs bad with linear classifiers since they are trained without dropout and cannot cope well with missing inputs. \\
TypiClust is better with embedded data not only due to lower budgets. On other datasets it is not able to outperform other algorithms in early stages

@@

\section{Comparison of Different Classifier Sizes}
We tested two different classifier sizes in Splice and DNA:
\begin{itemize}
	\item Small: [24, 12] (2400 parameters)
	\item Big: [24, 48, 48] (5700 parameters)
\end{itemize}
\begin{figure}
	\centering
	\includegraphics[width=0.49\linewidth]{img/supp_dna_small.png}
	\includegraphics[width=0.49\linewidth]{img/supp_dna_big.png}
	\includegraphics[width=0.49\linewidth]{img/supp_micro_dna_small.png}
	\includegraphics[width=0.49\linewidth]{img/supp_micro_dna_big.png}
	\caption{Comparison of small and big classifiers for the DNA dataset}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.49\linewidth]{img/supp_splice_small.png}
	\includegraphics[width=0.49\linewidth]{img/supp_splice_big.png}
	\includegraphics[width=0.49\linewidth]{img/supp_micro_splice_small.png}
	\includegraphics[width=0.49\linewidth]{img/supp_micro_splice_big.png}
	\caption{Comparison of small and big classifiers for the Splice dataset}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Choosing the Classifier}\label{sec:choosing_the_classifier}
Traditionally, the classifier is chosen per dataset so that it is capable of solving the dataset close to the SOTA performance reported in the literature.
Since we are not interested in archiving a new SOTA in any classification problem, we opt to use smaller classifiers for the following reasons:
Smaller classifiers generally (i) exhibit more stable training behavior and (ii) on average require less sampled datapoints to reach the their upper bound performance on the full dataset.
For every dataset the chosen architecture's hyperparameters are optimized by to archive maximum upper bound performance.
One desired characteristic of small classifiers is that the ranking of AL algorithms should stay the same when switching to larger models.
A small analysis of this behavior can be found in Appendix \ref{app:classifier_size}.
We found that the ranking of AL algorithms unfortunately does change, but we did not observe systematics that benefit one or few specific algorithms.
We therefore rely on the different data domains to provide classification models of different sizes and archetypes to cover all of the use-cases.
For an overview of architectures and hyperparameters please refer to Appendix \ref{app:hyperparameters}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Training the Classifier}\label{sec:training_the_classifier}
The classification model can be trained in two ways. Either you reset the parameters after each AL iteration and train the classifier from scratch with the updated labeled set $\mathcal{L}^{(i)}$, or you retain the previous state and only fine-tune the classifier on $\mathcal{L}^{(i)}$ for a reduced number of epochs.
In this work we use the fine-tuning method for raw datasets to save computation, while we use the from-scratch training for embedded dataset, since they have very small classifiers and this method generally produces better results.
Our fine-tuning scheme always trains for at least one epoch and employs an aggressive early stopping after that.
The early stopping has patience 0, so it will stop as soon as the validation loss does no longer decrease.
Even though the use of a fully labeled validation set might be regarded as impractical, since such a set will never exist during deployment, we strongly advocate for using it in benchmarks to control the classifier training.
In this work we use the validation set to optimize the hyperparameters of the classifier and reduce overfitting with early stopping the training process in every iteration.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}\label{sec:datasets}
For all our datasets we use the pre-defined train / test splits, if given. 
In the remaining cases, we define test sets upfront and store them into separate files to keep them fixed across all experiments.
The validation set is split during experiment-time and depends on the dataset-seed.\\
\textbf{Tabular:}
We use \textbf{Splice}, \textbf{DNA} and \textbf{USPS} from LibSVMTools \cite{libsvmtools}.
All three datasets are normalized between [0, 1]. \\
\textbf{Image:}
We use \textbf{FashionMNIST} \cite{xiao2017fashion} and \textbf{Cifar10} \cite{krizhevsky2009learning}.
Both datasets are normalized according to their standard protocols. \\
\textbf{Text:}
We use \textbf{News Category} \cite{misra2022news} and \textbf{TopV2} \cite{chen-etal-2020-low-resource}.
For News Category we use  the 15 most common categories as indicated by its Kaggle site.
We additionally drop sentences above 80 words to reduce the padding needed (retaining 99,86\% of the data).
For TopV2, we are only using the "alarm" domain.
Both datasets are encoded with pre-trained GloVe (Common Crawl 840B Tokens) embeddings \cite{pennington2014glove}.
Since neither dataset provided a fixed test set, we randomly split 7000 datapoints into a test set. \\ [1mm]
%
We would like to point out that these datasets can be considered "toy-datasets" and therefore not relevant for practical purposes.
This might be true if we aimed to develop novel classification models on these datasets, however, similar to our argumentation for picking smaller classifiers, we are solely focused on comparing different AL algorithms in this paper.
Our core assumption is that a well-performing algorithm in our benchmark will also transfer into more practical use-cases. \\ [1mm]
Adapting the experimental setting from \cite{hacohen2022active} we offer all our datasets in the raw setting as well as pre-encoded by a fixed embedding model that was trained by unsupervised contrastive learning. 
The text datasets are an exception, as they are only offered in their encoded form.
The pre-encoded datasets enable us to test our single-sample algorithms on more complex datasets like Cifar10 and FashionMnist without the need of sampling $>2000$ datapoints before we can reach our upper bound performance.
The embedding model was trained with the SimCLR \cite{chen2020simple} algorithm. 
For Cifar10 we adapt the reported hyperparameters from \cite{hacohen2022active} and for the tabular datasets we use random search to optimize the hyperparameters.
The quality of embeddings during pretext training was measured after each epoch by attaching a linear classification head and evaluating this classifier for test accuracy, mirroring our AL setup for embedded datasets.

@@

\begin{table}[]
	\centering
	\begin{tabular}{l|lll}
		& Splice        & DNA           & USPS          \\
		\hline
		Oracle          & 0.830 +- 0.01 & 0.836 +- 0.02 & 0.823 +- 0.01 \\
		SAL & 0.799 +- 0.01 & 0.797 +- 0.03 & 0.809 +- 0.01 \\
		Coreset & 0.800 +- 0.01 & 0.795 +- 0.03 & 0.787 +- 0.02 \\
		TypiClust       & 0.790 +- 0.01 & 0.771 +- 0.04 & 0.761 +- 0.02 \\
		MarginScore     & 0.797 +- 0.02 & 0.795 +- 0.04 & 0.808 +- 0.01 \\
		ShannonEntropy  & 0.799 +- 0.02 & 0.794 +- 0.04 & 0.807 +- 0.01 \\
		RandomAgent     & 0.788 +- 0.01 & 0.765 +- 0.03 & 0.772 +- 0.01 \\
		Badge           & 0.807 +- 0.01 & 0.769 +- 0.06 & 0.797 +- 0.02 \\
		BALD            & 0.811 +- 0.01 & 0.743 +- 0.04 & 0.717 +- 0.05
	\end{tabular}
	\caption{AUC values for all algorithms on the tabular datasets. Higher is better.
	{\color{red}Is the STD even helping us here? Since we repeated 50 times, we can use the values as "true medians"} }
\end{table}
%
\begin{figure}
	\centering
	\includegraphics[width=0.49\linewidth]{img/eval_splice}
	\hspace{1mm}
	\includegraphics[width=0.49\linewidth]{img/eval_dna}
	\caption{Results for all algorithms on Splice and DNA, both from the tabular domain. Even within one domain, the performance of the same algorithm can vary drastically.}
	\label{fig:eval_vector}
\end{figure}
%
\begin{figure}
\centering
	\includegraphics[width=0.49\linewidth]{img/vector.png}
	\includegraphics[width=0.49\linewidth]{img/vector_enc.png}
	\includegraphics[width=0.49\linewidth]{img/text.png}
	\includegraphics[width=0.49\linewidth]{img/all_data.png}
	\caption{Critical Difference Diagram for all algorithms grouped by domain and all domains combined. Ranks are computed based on median AUC for each algorithm and dataset combination. Lower ranks are better.}
	\label{fig:cd_diagrams}
\end{figure}
%
From Fig. \ref{fig:eval_vector} we notice drastically different qualities for the same AL algorithm for different datasets.
We would like to highlight that both datasets are tabular from the medical domain with similar number of features and classes, yet we see that i.e. BALD is the best algorithm for Splice and the worst algorithm for DNA.
These inconsistencies are present between the datasets of all our tested domains, further highlighting the difficulties for comparing AL algorithms in terms of average performance.
In order to provide a meaningful analysis of which algorithm can be expected to perform best on average we ranked the algorithm for each dataset based on their median AUC and displayed these rankings in critical difference diagrams \cite{IsmailFawaz2018deep}.
In Fig. \ref{fig:cd_diagrams} we report the rankings split by domain as well as across all domains (excluding the toydata). \\ [2mm]
%
BALD performs bad with linear classifiers since they are trained without dropout and cannot cope well with missing inputs. \\
TypiClust is better with embedded data not only due to lower budgets. On other datasets it is not able to outperform other algorithms in early stages

@@

\section{Comparison of Different Classifier Sizes}\label{app:classifier_size}
We tested two different classifier sizes in Splice and DNA:
\begin{itemize}
	\item Small: [24, 12] (2400 parameters)
	\item Big: [24, 48, 48] (5700 parameters)
\end{itemize}
\begin{figure}
	\centering
	\includegraphics[width=0.49\linewidth]{img/supp_dna_small.png}
	\includegraphics[width=0.49\linewidth]{img/supp_dna_big.png}
	\includegraphics[width=0.49\linewidth]{img/supp_micro_dna_small.png}
	\includegraphics[width=0.49\linewidth]{img/supp_micro_dna_big.png}
	\caption{Comparison of small and big classifiers for the DNA dataset}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.49\linewidth]{img/supp_splice_small.png}
	\includegraphics[width=0.49\linewidth]{img/supp_splice_big.png}
	\includegraphics[width=0.49\linewidth]{img/supp_micro_splice_small.png}
	\includegraphics[width=0.49\linewidth]{img/supp_micro_splice_big.png}
	\caption{Comparison of small and big classifiers for the Splice dataset}
\end{figure}
